### 2022-11-23 Manuscripts/cogsci-journal/revision/supp.tex

%!TEX output_directory = latex_out/
\documentclass[12pt]{article}
\usepackage[letterpaper, margin=1in, headheight=15pt]{geometry}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{pgfplots}
\usepackage{fancyhdr}
\usepackage{csquotes}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage[toc]{appendix}
\usepackage[natbibapa]{apacite}

@@

% set up notes-- different backgrounds for Nolan and Joe!
\newcommand\nbcnote[1]{\todo[inline, backgroundcolor = yellow]{\textbf{NBC}: #1}}
\newcommand\jlanote[1]{\todo[inline, backgroundcolor = lime]{\textbf{JLA}: #1}}
\newcommand\sxlnote[1]{\todo[inline, backgroundcolor = cyan]{\textbf{SXL}: #1}}

% set up header % 
\pagestyle{fancy}
\fancyhf{} % sets both header and footer to nothing
\renewcommand{\headrulewidth}{0pt}
\lhead{RUNNING HEAD: Similarity, Contrast, and Representativeness in Categorization}
\fancyhead[R]{\thepage}

\begin{document}

% ------- TITLE PAGE ------- %
\begin{center}
\hfill
\\[1in]

Creating Something Different: \\
Similarity, Contrast, and Representativeness in Categorization

\vfill
Joseph L. Austerweil\textsuperscript{1}
Nolan Conaway\textsuperscript{1}, 
Shi Xian Liew\textsuperscript{1},
and Kenneth J. Kurtz\textsuperscript{2}
\\[\baselineskip]
\textsuperscript{1}Department of Psychology, University of Wisconsin-Madison, Madison, WI, USA
\textsuperscript{2}Department of Psychology, Binghamton University, Binghamton, NY, USA
\\[1in]

\vfill

Author Note

Correspondence concerning this article should be addressed to: 
Joseph L. Austerweil, 1202 West Johnson Street, Madison, WI 53706.
E-mail: austerweil@wisc.edu

\end{center}
\clearpage

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking whether (and if so, how) contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.

\setlength\parindent{0.5in}
{\em Keywords}: categorization; concepts; category learning; generation;
computational modeling; exemplar models; representativeness
\clearpage

\begin{flushleft}

\section{Introduction}
\setlength\parindent{0.5in}

Creating new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. Computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create 
new categories that are as interesting as cell phones and smart phones? Devising such an experiment is still beyond our capability. Instead, we take a step towards this goal, and investigate the fundamental constraints and expectations that people have when they generate new categories.

% we do have tools to study the fundamental constraints on category . As a
%step towards these goals, in this article we investigate the expectations that
%people have when they generate new categories using computational
%models and behavioral experiments.

Creating new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect human learning of an object's
category given its features \citep{shepard1961learning,kurtz2013human}, or the
prediction of an object's unobserved features given some of its other features
and/or its category \citep{markman2003category}. The generation of members of a
new category consists of inferring {\em all} features for a {\em novel} category
label. Thus, we can make progress formalizing the processes involved in category
generation by extending theories of categorization to this case. This will also serve to test the breadth of the explanatory power of categorization models to tasks beyond typical category learning.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Contrast from other
known categories should play a role. Indeed, scientists in other fields, such
as marketing \citep{berger16} and sociology \citep{rogers03}, highlight the
critical importance of contrast in the creation of new ideas.

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Intuitively, the influence of contrast may be expected in a number of
real-life scenarios involving the generation of new categories. For instance, a
college student who has been eating instant noodles for a week may be more
inclined to seek out something healthier and less carbohydrate-heavy (e.g., a
Caesar salad) compared to something similar (a different flavour of instant
noodles). An actor who is dissatisfied with being typecast as a villain may seek
out roles as heroes as opposed to playing other antagonistic characters. An
especially compelling example can be seen in \cite{askin2017} who found that
new songs that were more different to their peers were generally more popular
than new songs that were similar to others. It appears that once there is the
motivation to generate something new, there is an associated motivation to
generate something \emph{different}.

Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is discussed
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another.

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic \citep{kahneman1972subjective}
states that people make judgments regarding an outcome based on how
representative it is of the evidence given in the current context. The heuristic
is a powerful theoretical construct that has been used to capture a range of
complex patterns of human judgments \citep{kahneman73,tversky74,tversky83},
especially those that deviate from normative theory (as given by a
straightforward application of Bayes' rule). For example, the coinflip sequence
TTHTH is perceived to be more random (due to it being representative of a
coinflip sequence) than TTTTT (despite both having the same probability of
having been generated from a fair coin). Although there is a healthy debate as
to whether perceived randomness {\em really} deviates from normative theory
\citep{griffiths18,hahn09}, the representativeness heuristic remains one of the
main explanations of human judgments \citep{reimers18}. To formalize the
hypothesis that people expect a new category to be representative of the
opposite of the current category, or contrast as {\em representativeness}, we
use the Bayesian formulation of representativeness \citep{tenenbaum01}, which
has been used successfully to capture human performance in color categorization
\citep{abbott16}, image category learning \citep{abbott11}, and language grammar
learning \citep{rafferty10}. We do so by extending the state-of-the-art category
generation model \citep{jern2013probabilistic} and find that human category
generation is best captured as generating representative, rather than probable,
samples.

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the behavioral results and the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

\section{Prior work}

Much of what we know about concept generation and contrast comes from the
foundational literature on creative cognition. In a series of reports, Ward and
colleagues
\citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining}
established that category generation is highly constrained by prior knowledge:
Generated categories tend to consist of features observed in known categories,
and they tend to exhibit the distributional properties found in known
categories. In a seminal study, \cite{ward1994structured} asked participants to
generate new species of alien animals by drawing and describing members of the
species. People tended to generate species with the same features as on Earth
(e.g., eyes, legs, wings), and possessing the same feature correlations as on
Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same
species tended to share more features with one another compared to members of
opposite species.

The broader set of observations made by Ward and colleagues provide a great deal
of insight into the role of prior knowledge in constraining category generation.
Much of the work from this area
\citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how
information provided to participants (such as an example of a species generated
by other participants) can drastically diminish the difference of a new category
from pre-existing categories. Theoretical accounts of these effects have
primarily been grounded within the categorization literature. For example, the
predominant ``Path of Least Resistance'' account
\citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when
generating a new species of animal, people retrieve from memory a known
subcategory of animals (e.g., {\em bird}, {\em dog}, {\em horse}), and simply
change some of the features to make something new. People are thought to change
only features that are not characteristic of the retrieved category (e.g., if
{\em bird} was retrieved, the presence of {\em wings} would not change, but {\em
  color} might). This theory incorporates elements of the highly influential
basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as
well as the exemplar view \citep{medin1978context,brooks1978nonanalytic}. While
this work has been incredibly useful in providing a conceptual sketch of
generation theories, its qualitative nature and the hand-drawn responses used in
the experiments paradigms precludes the development of formal approaches that
can be used to test them in a fine-grained manner.

\cite{jern2013probabilistic} recently showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after learning categories with a positive correlation between the size
and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. By
replicating \cite{ward1994structured}, they demonstrated that category generation
can be studied in a well-known and highly controlled experimental paradigm.

The authors evaluated the predictions of several formal models on their data.
Most notably, they showed that a hierarchical Bayesian model provided the
strongest account of their results. Their model views observed examples as
samples from an underlying category distribution, describing the location of the
category in the space, as well as how it varies along each feature. In turn,
each category is viewed as a sample from an underlying {\em domain}
distribution, specifying distributional commonalities among the observed
categories. Generated categories are thought to stem from the same domain
distribution as observed categories, thus the distributional properties of
observed categories will be preserved within the generated category.

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity to which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of their results, as the
experiments devised by \citeauthor{jern2013probabilistic} were specifically
designed to challenge it.

It is worth pointing out that the early observation of regularities in
distributional properties across categories is not confined to the work of Ward
and colleagues. Most notably, \citet{thomas98} trained participants to classify
circles of different sizes, each with a radial line of varying orientation, into
two categories. During a surprise prediction phase, participants were asked to
generate a value for a missing feature for a certain category exemplar given a
value on the other feature. Their results revealed that exemplars tended to
share the same feature correlations as exemplars from previously learned
categories. Interestingly, this effect was not consistently observed when the
given values of a feature fell outside the range of learned values for that
particular category. For instance, if the learned Category A exemplars were
generally 8 cm to 10 cm in radius and negatively correlated with the angle of
the radial line, participants generally produced radial line orientations that
were also negatively correlated with the exemplar size when the given size was
between 8 cm and 10 cm. However, when the given size fell outside that range for
that category (e.g., 14 cm for a Category A exemplar), some participants
consistently produced radial line orientations that were not negatively
correlated with the size and that also placed the exemplars in a significantly
different location in the feature space. These results suggest that when tasked
to produce exemplars that fall outside what was previously learned, people show
some tendency to generate exemplars that are different (both distributionally
and spatially) to what was learned. We explore this idea of contrast
more deeply in the following section.

\section{And Now for Something (Completely) Different: The Role of Contrast}

%Prior work on category generation has explored only one factor: distributional
%correspondences between learned and generated categories. As a result, most of
%the computational, theoretical, and empirical efforts have been focused on
%explaining those effects. In this paper, we investigate another important
%factor: category contrast. To generate a novel concept, individuals must produce
%something that is in some capacity {\em different} from what they already know.
%As a consequence, we propose that contrast should be a primary constraint that
%underlies people's expectations about how a set of categories are organized:
%Categories should be different from each other.

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over stimulus space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

The copy-and-tweak model tested by \cite{jern2013probabilistic} also claims
little about how generated categories should contrast with what is already
known. In their simulations, the model was only tested on generation after the
learner had been exposed to members of the target category, and so the model's
ability to generate a new category from scratch was not evaluated. However, the
model's generation is based exclusively on similarity to known members of the
{\em target} category; when there are no members of the target category,
generation is presumably random.

\subsection{Contrast as exemplar dissimilarity: The PACKER Model}

\label{section:PACKER-definition}

As noted above, the constraint that new concepts should differ from what is
already known has been largely overlooked in previous work. This is no doubt in
part due to the vague definition of what it means for a concept to be
``different'': A generated category may be different from what is already known
in any number of respects. Towards providing a more precise definition of the
role of contrast in generation, we formalized contrast in a novel exemplar
model, PACKER ({\em Producing Alike and Contrasting Knowledge using Exemplar
  Representations}). PACKER explains category generation as a balance between
two fundamental constraints: The category to be generated should not be similar
to known categories, and exemplars within each category should be similar to one
another. These ideas are implemented within the well-studied exemplar framework
-- the PACKER model is an extension of the influential Generalized Context Model
of categorization \citep[GCM;][]{nosofsky1984choice,nosofsky1986attention}.

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been throughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns how contrast may be captured through the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

Both PACKER and the GCM simulate categorization under the assumption that
learners represent categories as a collection of exemplars, corresponding to the
labeled stimuli they have observed. The exemplars are encoded within a
$k$-dimensional psychological space, and model performance is based on the
amount of similarity between the item to be categorized and the stored
exemplars. Similarity between two examples, $s\left(x_i, x_j\right)$, is
computed as an inverse exponential function of distance
\citep[following][]{attneave1950,shepard1957stimulus,shepard1987toward}:

\begin{equation} s\left(x_i,x_j\right) = \exp \left\{ -c \left[\sum_{k}{ w_k
\left| x_{ik} - x_{jk} \right|^r }\right]^{1/r} \right\}
\label{eq:similarity}
\end{equation}
% 
where $w_k$ is the attention weighting of dimension $k$ ($w_k \geq 0$ and
$\sum_k{w_k} = 1$), accounting for the relative importance of each dimension in
similarity calculations, and $c$ ($c>0$) is a specificity parameter controlling
the spread of exemplar generalization. For simplicity, attention will be
distributed uniformly in our simulations (unless otherwise noted). The value of
$r$ depends on the nature of the experimental conditions being simulated: $r=1$
is appropriate for separable dimensions, whereas $r=2$ is appropriate for
integral dimensions \citep[e.g.,][]{shepard1964attention,garner1974processing}.
In our simulations, we set $r=1$ due to the separable nature of the stimulus
dimensions used in our experiments (see Figure \ref{fig:e1-conditions}).

PACKER (as well as its name) was in part inspired by earlier work from the
categorization literature \citep{hidaka2011packing,stewart2005}. \citet{hidaka2011packing} argued that
natural categories ``pack'' the values of features such that different
categories fill the domain space with distance between one another, while
keeping items within the same category close together. Inspired by this idea,
PACKER proposes that generation is constrained by both similarity to members of
the target category (the category in which a stimulus is being generated) as
well as similarity to members of other categories: the most desirable generation
candidates are similar to members of the target category and not similar to
members of contrast categories. This is achieved by aggregating similarity
across known exemplars differently according to class membership. The aggregated
similarity $a(y,x)$ between generation candidate $y$ and stored exemplars $x$ is
given by:

\begin{equation} a(y, x) = \sum_j{f(x_j) s(y, x_j)}
\end{equation}
% 
where $f(x_j)$ is a function specifying each exemplar's contribution to
generation. A negative value for $f(x_j)$ produces a `repelling' effect (items
are less likely to be generated nearby $x_j$), and a positive value produces an
`attracting' effect (items are more likely to be generated nearby $x_j$). When
$f(x_j)=0$, the exemplar does not contribute to generation.

PACKER sets $f(x_j)$ depending on exemplar $x_j$'s category membership:
$f(x_j) = \theta_t$ if $x_j$ is a member of the target category, and
$f(x_j) = -\theta_c$ if $x_j$ is a member of a contrast category. $\theta_t$ and
$\theta_c$ are free parameters ($0 \leq \theta_t,\theta_c$) controlling the
trade-off between within- and between-category similarity. For example, when
$\theta_t = \theta_c = 0.5$, $f(x_j) = 0.5$ for members of the target category
and $f(x_j) = -0.5$ for members of other categories; thus, the model is likely
to generate items that are similar to members of the target category but are not
similar to members of other categories. In this way, $\theta_t > 0$ with
$\theta_c = 0$ produces exclusive consideration of target-category members, and
$\theta_c > 0$ with $\theta_t = 0$ produces exclusive consideration of
contrast-category members. The combination of $\theta_t$ and $\theta_c$
parameters thus specifies a wide breadth of possible approaches; by fitting it
to a dataset, one can describe the relative roles of between-category contrast
and within-category similarity in generation. See Figure
\ref{fig:packer-examples} for an illustration of how these parameters control
the relative influence of within-category similarity and contrast to other
categories when generating a new exemplar.

\begin{figure}
	\begin{center} \inputpgf{figs/}{packer-examples.pgf}
      \caption{PACKER generation of a category `B' example, following exposure
        to one member of category `A' and one member of category `B'.
        Predictions are shown for three different parameterizations (differing
        in values of $\theta_t$ and $\theta_c$): {\em (a)}
        Predictions based on contrast similarity only. {\em (b)}
        Predictions based on target similarity only. {\em (c)} Predictions with
        both constraints considered.}
		\label{fig:packer-examples}
	\end{center}
\end{figure}

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.
It is worth noting that PACKER is only one possible exemplar-based account of
category generation within our proposed framework. That is, PACKER places
specific constraints on the possible values of $f(x_j)$, but other
exemplar-based category generation models with drastically different behavior
can be formalized in this framework by imposing alternative constraints. For
example, as will be discussed in more detail below, PACKER is formally
equivalent to the copy-and-tweak model proposed by \cite{jern2013probabilistic}
when $\theta_c = 0$ and $\theta_t > 0$. Likewise, when
$\theta_t = 0$ and $\theta_c > 0$, PACKER can represent a
contrast-only generation mode, relying exclusively on contrast when generating
new categories. When $f(x_j) < 0$ for all $x_j$ (regardless of class
membership), a ``pure-packing'' approach is yielded, generating items in
unoccupied areas of the domain. Thus, the proposed framework may be used to
describe a wide variety of qualitatively distinct generation strategies.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
%: Both models are
%exemplar-based, and both models generate new items according to their similarity
%to known members of the target class. However, PACKER diverges from the
%copy-and-tweak model by including a contrast mechanism, enabling generation
%according to dissimilarity to members of opposing categories. As a consequence,
%copy-and-tweak can be realized as a parameterization of the PACKER model that is
%insensitive to category contrast. Specifically, when $\theta_c = 0$ and
%$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
%$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
%items, and is mathematically equivalent to a copy-and-tweak approach.

\label{section:copytweak-packer} The PACKER model is similar to the copy-and-tweak model reported by \cite{jern2013probabilistic}.
An advantage of formulating copy-and-tweak as a special case parameterization of PACKER. In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and thus, we
leave extending PACKER to incorporate distributional factors for future work.

\subsection{Contrast as representativeness: An extension to Jern and Kemp
(2013)}
\label{section:representativeness-definition}

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

The Bayesian formalization of representativeness makes it straightforward to
extend the existing hierarchical Bayesian model of category generation
developed by \cite{jern2013probabilistic}.\footnote{We thank Charles Kemp for the suggestion.} Specifically, both models assume that
exemplars are sampled from a given category distribution ($h$ in
Equation \ref{eq:representativeness}). Each category distribution is a
multivariate normal parameterized by a location vector $\mu_k$ and covariance
matrix $\Sigma_k$ (i.e., $h = \mathcal{N}(\mu_k,\Sigma_k)$). The parameters
$\mu_k$ and $\Sigma_k$ are assumed to be samples from a prior
normal-inverse-Wishart (NIW) distribution parameterized by
$\mu_0,\Sigma_0,\kappa,\nu$. Mathematically,
\begin{align}
  \mu_k, \Sigma_k|\mu_0,\Sigma_0,\kappa,\nu &\sim
                                              \textrm{NIW}(\mu_0,\Sigma_0,\kappa,\nu)
                                              \label{eq:mvnsample}\\
  y|C &\sim \mathcal{N}(\mu_k,\Sigma_k) \label{eq:exemplarsample}
\end{align}

In our simulations, we set the prior mean $\mu_0$ to the center of the stimulus
space and the prior variance to be isotropic ($\Sigma_0 = \lambda \mathbf{I}$ where
$\lambda$ is a free parameter and $\mathbf{I}$ is a $d$-by-$d$ identity matrix
with $d$ representing the number of dimensions or features in the domain). The
$\kappa$ and $\nu$ parameters are freely estimated within the constraints
$\kappa > d-1$ and $\nu > 0$.

With the assumption of the NIW prior, the expected location vector $\mu_k$ given exemplars observed from category $k$ is:
\begin{equation}
  \mu_k = \dfrac
    {\kappa\mu_{0} + n_k \bar{x}_k}
    {\kappa + n_k}
    \label{eq:category_mus}
\end{equation}
where $n_k$ is the number of observed exemplars in category $k$ and $\bar{x}_k$
is the observed category mean. Note that if there are no observed exemplars in
the target category (i.e., if the model is generating a completely novel
category), Equation \ref{eq:category_mus} simplifies to $\mu_k = \mu_0$.

The NIW prior also allows us to infer the category covariance matrix $\Sigma_k$
by computing the following:
\begin{equation}
  \Sigma_k = [\nu \Sigma_D+ C_k +
    \dfrac
    {\kappa n_k}
    {\kappa + n_k}
    (\bar{x_k}-\mu_k)(\bar{x_k}-\mu_k)^T
  ] (\nu + n_k)^{-1}
  \label{eq:Sigma_y}
\end{equation}
where $C_k$ is the observed category covariance and $\Sigma_D$ is the domain
covariance matrix from which $\Sigma_k$ samples are obtained. We can
infer $\Sigma_D$ from the observed category covariances $C$ and the prior
covariance $\Sigma_0$. Specifically,
\begin{equation}
    \Sigma_D = \Sigma_0 + \sum_{k}{C_k}
\end{equation}

At this point, both the representativeness model and the hierarchical Bayesian
model from \cite{jern2013probabilistic} are largely identical. However, both
models diverge in their computation of exemplar generation probabilities. While
the original hierarchical Bayesian model produces novel exemplars with
probabilities proportional to the multivariate normal likelihood $p(y|h)$, the
representativeness model generates new exemplars according to their representativeness $R(y,h)$ as formulated in Equation \ref{eq:representativeness}. In practice, the probability of generating a
particular candidate $y$ is obtained using an Exponentiated
choice rule \citep{luce1977choice}:
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot R(y,h))}
    {\sum_k \exp(\theta  \cdot R(y_k,h_k))} .
\label{eq:representative-choice}
\end{equation}
where $\theta$ is a freely estimated response determinism parameter (constrained
such that $\theta \geq 0$).

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
with the assumption of unimodal underlying distributions, this mechanism of
representativeness results in the generation of novel exemplars that are less
similar to exemplars from the known category. This occurs because the
representativeness of novel exemplars tend to be highest where the probability
density of the known category's underlying distribution is the lowest. We see
this illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.
Exceptions to this pattern can occur (i.e., the generation of novel exemplars
similar to known category exemplars) when the known category distribution is
especially large on all features -- we explore this in more detail in Section
\ref{section:diffcon} where we highlight an important condition demonstrating
how PACKER and the representativeness model can make qualitatively different
predictions.

\begin{figure}
	\begin{center} \inputpgf{figs/}{hbm-examples.pgf}
		\caption{Generation of a category `B' example, following exposure
to one member of category `A' and one member of category `B'. Predictions are
shown for (a) the original hierarchical Bayesian model and (b) the
representativeness model.}
		\label{fig:representative-examples}
	\end{center}
\end{figure}

%\subsection{Synopsis and Prognosis}
%
%Research on categorization and category generation has focused on the finding
%that generated categories tend to possess distributional commonalities with
%known categories. However, a fundamental task of categorization is to create
%something {\em new} (i.e., different from what is already known). The manner in
%which generated categories differ from known ones is, nonetheless, poorly
%understood: Existing theories do not make strong predictions about how generated
%concepts should systematically differ from existing ones. Above, we introduced
%two novel approaches to formalizing the roles of similarity and contrast in
%category generation.
%
%In the sections below, we present two experiments demonstrating systematic
%effects of category contrast on category generation inspired by factors
%influencing how the contrast models generate new categories. Our experiments are
%based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward
%translation of the traditional artificial classification paradigm to the task of
%category generation: Participants are first exposed to a single,
%experimenter-defined category, and are then asked to generate members of a new
%category. We then report quantitative analyses of our results comparing the exemplar dissimilarity
%and the representativeness accounts of our results to that of the hierarchical
%Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.
%After analyzing the category generation studies, we present a third experiment to
%distinguish between our two conceptualizations of contrast.

\section{Experiment 1}
To begin our investigation, we examined category generation in a well-understood domain using a few disparate category types. We used an
artificial stimulus design: A two dimensional domain of squares, varying in
color and size (see Figure \ref{fig:e1-conditions}a). These dimensions
have been used in numerous classification learning studies
\cite[e.g.,][]{conaway2016similar,conaway2016generalization,shepard1961learning,nosofsky1994comparing}.
Unlike those used in the \cite{jern2013probabilistic} experiments, distance on
these physical dimensions aligns more directly with perceptual similarity,
allowing us to evaluate the role of  contrast in categorization
more precisely. It also enables more straightforward comparisons to prior work. We tested the effects of category contrast after learning one category from a set of qualitatively distinct category structures, as shown in Figure \ref{fig:e1-conditions}.

Figures \ref{fig:e1-conditions}b-d show the values of exemplar dimensions
belonging to the experimenter-defined categories (`A', or `Alpha') that
participants were assigned to learn about prior to generating a new category.
Each participant learned one of the category types during training. In the
`Cluster' type, category A is a tight cluster of examples in the space.
Perceptually instantiated, the members of category A might, for example, be
large and dark in color. In the `Row' type, category A has a row pattern across
the space, varying along one feature but not the other. Thus, its members might
all be dark in color but would vary in size. Finally, in the `XOR' type, the
experimenter-defined category consists of two clusters separated in opposite
corners of the space, conforming to the exclusive-or logical structure (e.g.,
members are small and dark or large and light).

It should be noted that in our experiments the assignment of the perceptual
to conceptual dimensions (e.g., $X \rightarrow Size$, $Y \rightarrow Color$) and the direction of variation along each dimension (e.g., $dark
\rightarrow light$ or $light \rightarrow dark$) were counterbalanced across
participants. The category types in Figure \ref{fig:e1-conditions} are plotted
in conceptual space, rather than perceptual space. Thus, while the
conceptual organization of the category types remains constant, each category
type may have a different physical instantiation according to the counterbalance
assignment. For example, the Cluster type may be large and dark in color, or it
may be small and light in color, depending on the assignment and direction of
the dimensions. For this reason, below we will discuss generation within a
conceptual space, rather than a physically instantiated one.

>>>>

%!TEX output_directory = latex_out/

\documentclass[12pt]{article}
\usepackage[letterpaper, margin=1in, headheight=15pt]{geometry}
\usepackage{amsmath,fancyvrb}
\usepackage[natbibapa]{apacite}
\usepackage{pgfplots}
\usepackage{setspace}

@@

\begin{document}
\doublespacing
\section{Finding Contrast Effects in Prior Work}
